Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-12 aistats.tex
Add a conclusion

\begin{table}[t]
\begingroup
\renewcommand*{\arraystretch}{1.25}%
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\centering
\begin{tabular}{lccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Variance on all $\Theta$ % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark  & \redxmark
\\
Variance at $\theta_*$ %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark  & \greencmark
\\
Optimality gap %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\caption{
Summary of results about SMD with relative smoothness and relative strong convexity assumptions. 
Barrier tells whether the assumption can hold with barrier objectives.
$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
}
\label{tbl:assumptions}
\endgroup
\end{table}

\subsubsection{Bounded Variance}
\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
\alignn{
\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
\label{eq:assumption-hanzely}
}
for some constant $C$.
For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the LHS of \cref{eq:assumption-hanzely} is bounded by
$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
to hold.

Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.

@@

\citet{dragomir2021fast} introduce a much weaker assumption $\forall n \geq 1, \forall \MAPm,$
\alignn{
	\expect[\tilde g]{
		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
	} \leq 2 \lr^2 C \; .
	\label{eq:assumption-dragomir}
}
In words, this is bounding the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality. 
Under this assumption, they prove convergence up to a variance ball with a constant step-size.
Their descent lemma \citep[Eq.(12)]{dragomir2021fast} is the exact analog  of a modern analysis of SGD \citep[Th.3.2]{gower2019sgd}, meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.

@@

\subsubsection{Bounded Optimality Gap}
Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explores the hypothesis
\aligns{
\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C 
}
where $f_X$ is a stochastic estimate of $f = \expecti{f_X}$. In our case $f_X(\nat) = - \log p(X\cond \nat)$.
In words, this is lower bounding the expectation of the minimum of these stochastic estimates.
For probabilistic models, such a bound is finite as soon as the model cannot give infinite density to any data point $x$.
This holds for instance for discrete distributions 
-- $f_x\geq 0$ because the probability mass is upper bounded by $1$ -- 
but it rules out  many families.
For normal distributions set $\mu=x$ and $\sigma^2 \rightarrow 0$ to get $p(x\cond \nat) \rightarrow +\infty$. 
For gamma, set $\alpha = \beta x$ and $beta \rightarrow +\infty$.
For beta, set $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
 Other examples include inverse Gaussians, log-normal, gamma, inverse gamma.

@@

If assumptions~\eqref{eq:assumption-hanzely} and~\eqref{eq:assumption-dragomir} do not hold for every iterate, then let's assume they hold on average. Does the proof remain valid ? Yes I think so \TODO{check this}. Does it become impossible to prove / as hard as the initial problem ? Yes.
\TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}

>>>>

\begin{table}[t]
\begingroup
\renewcommand*{\arraystretch}{1.25}%
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\centering
\begin{tabular}{lccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Variance $\forall \nat\in\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark  & \redxmark
\\
Variance at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark  & \greencmark
\\
Optimality gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\caption{
Summary of results about SMD with relative smoothness and relative strong convexity assumptions. 
Barrier tells whether the assumption can hold with barrier objectives.
$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
}
\label{tbl:assumptions}
\endgroup
\end{table}

\subsubsection{Bounded Variance}
\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
\alignn{
\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
\label{eq:hanzely}
}
for some constant $C$.
For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the LHS of \cref{eq:hanzely} is bounded by
$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:hanzely}
to hold.
Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
\aligns{
\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
}
thus showing positivity.
Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.

@@

\citet{dragomir2021fast} introduce a much weaker assumption $\forall n \geq 1, \forall \MAPm,$
\alignn{
	\expect[\tilde g]{
		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
	} \leq 2 \lr^2 C \; .
	\label{eq:dragomir}
}
In words, this is bounding the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality. 
Under this assumption, they prove convergence up to a variance ball with a constant step-size.
Their descent lemma \citep[Eq.(12)]{dragomir2021fast} is the exact analog  of a modern analysis of SGD \citep[Th.3.2]{gower2019sgd}, meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.

@@

\subsubsection{Bounded Optimality Gap}
Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explores the hypothesis
\alignn{
\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C 
\label{eq:dorazio}
}
where $f_X$ is a stochastic estimate of $f = \expecti{f_X}$. In our case $f_X(\nat) = - \log p(X\cond \nat)$.
In words, this is lower bounding the expectation of the minimum of these stochastic estimates.
For probabilistic models, such a bound is finite as soon as the model cannot give infinite density to any data point $x$.
This holds for instance for discrete distributions 
-- $f_x\geq 0$ because the probability mass is upper bounded by $1$ -- 
but it rules out  many families.
For normal distributions set $\mu=x$ and $\sigma^2 \rightarrow 0$ to get $p(x\cond \nat) \rightarrow +\infty$. 
For gamma, set $\alpha = \beta x$ and $beta \rightarrow +\infty$.
For beta, set $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
 Other examples include inverse Gaussians, log-normal, gamma, inverse gamma.

@@

If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let's assume they hold on average. Does the proof remain valid ? Yes I think so \TODO{check this}. Does it become impossible to prove / as hard as the initial problem ? Yes.
\TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}

\section{Conclusion}
Exponential family MLE and MAP are used everywhere -- everytime someone computes a mean and variance to fit some bell-shaped data. 
MLE and MAP are both maximizing the (regularized) likelihood of the data, whose suboptimality is the Kullback-Leibler (KL) divergence.
The KL is a very natural measure of discrepancy between a model and the true disribution.
Yet we do not have convergence rates on the expected KL, even in the well-specified setting.
Such a rate could inform us on how many samples we may need, or on the influence of the prior.
There are many results close from such a rate, in particular large sample results \citep{kakade2010learning, ostrovskii2021finite}, which describe how many samples are needed to ensure a locally quadratic regime.
Another approach to obtain rates is stochastic optimization \citep{bach2013nonstronglyconvex}. 
Here we observed that MAP perfectly fits the framework of stochastic mirror descent with relative smoothness assumptions.
Yet none of the current analysis of SMD hold for the MAP, even on a simple family such as $\cN(0,\sigma^2)$, thus revealing area for progress in non-Euclidean optimization.
In writing this paper, we hope to attract attention to this fundamental problem, leading to progress in both optimization and statistics.
